---
title: "Meta-analysis-Bd-hypotheses"
author: "Molly"
date: "July 2021"
output:
  rmdformats::readthedown:
    highlight: tango
    code_folding: hide
  html_document:
    smart_extension: no
    toc: true
    toc_float: true
    toc_depth: 2
    df_print: paged
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```
# Setup & Data

## **Packages**
```{r}
library(tidyverse)
library(dmetar)
library(meta)
library(metafor)
```

## **Data Read-in**
```{r}

## FILTER OUT 0-0s!
MetaBdData = read_tsv("Meta_Bd_Data_July18.txt") %>%
  filter(Inclusion=="Yes")

MetaBdData_Genotype = MetaBdData %>%
  filter(Genotype.Analysis=="Y")%>%
  rename(mortality.endemic= mortality.local.or.endemic.Bd, Nendemic = n.local.or.endemic.Bd,Npandemic = n.nonlocal.or.pandemic.Bd, mortality.pandemic=mortality.nonlocal.or.pandemic.Bd)

MetaBdData_Location = MetaBdData %>%
  filter(Location.Analysis=="Y")%>%
  rename(mortality.localBd= mortality.local.or.endemic.Bd, nlocalBd = n.local.or.endemic.Bd,nnonlocalBd = n.nonlocal.or.pandemic.Bd, mortality.nonlocalBd=mortality.nonlocal.or.pandemic.Bd)

```

# **Historical Adaptation Hypothesis**

## Main effect model

```{r}

## using event rate data

## we need number of individuals data...recalculating percantages to numbers 
MetaBdData_Genotype2 = MetaBdData_Genotype  %>%
  mutate(Pandemic.ndead = round((mortality.pandemic-Amount.mortality)/100*Npandemic),
         Endemic.ndead=round((mortality.endemic-Amount.mortality)/100*Nendemic)) %>%
  mutate(Pandemic.nlive = Npandemic-Pandemic.ndead,
         Endemic.nlive = Nendemic-Endemic.ndead) %>%
  mutate(Endemic.ndead.cor = if_else(Endemic.ndead < 0, 0, Endemic.ndead))

## calculating the risk ratios from the raw # individuals data
Genotype_RR_data = metafor::escalc(measure="RR", ai=Pandemic.ndead, bi=Pandemic.nlive, ci=Endemic.ndead.cor, di=Endemic.nlive, data=MetaBdData_Genotype2, slab = paste(Study.Name))
## yi = log RR

## main model with a random term for study 
Genotype_basemod = rma.mv(yi=yi, V=vi,data = Genotype_RR_data, random = ~1|Shortcitation, slab = Genotype_RR_data$Study.Name,method = "REML")

## model results
summary(Genotype_basemod)

```

**Overall all risk ratio**

*estimate in model output is log RR*

```{r}
## converting log estimate to RR
exp(coef(Genotype_basemod)[[1]])

## https://www.metafor-project.org/doku.php/tips:meta_regression_with_log_rr

```

## Heterogeneity: I^2

```{r}
# from metafor package site
W <- diag(1/Genotype_RR_data$vi)
X <- model.matrix(Genotype_basemod)
P <- W - W %*% X %*% solve(t(X) %*% W %*% X) %*% t(X) %*% W
I2 = 100 * sum(Genotype_basemod$sigma2) / (sum(Genotype_basemod$sigma2) + (Genotype_basemod$k-Genotype_basemod$p)/sum(diag(P)))

I2

```

## Plots 

*now trying to using forest() because I started digesting the strange example code :-)*


```{r eval=FALSE, include=FALSE}

 Genotype_RR_data %>%
  ggplot(aes(x= Study.Name, y = exp(yi))) +
  geom_hline(yintercept=1, size = 1, color = "black", alpha = 0.4) +
  geom_point(aes()) +
  geom_errorbar(aes(ymin = exp(yi)-exp(vi), ymax = exp(yi)+exp(vi), alpha = 0.5), width = 0.2,show.legend = FALSE) + ## not sure about using vi here...???
  theme_classic()+
  scale_x_discrete()+
  xlab("Study")+
  ylab("Risk Ratio") +
  coord_flip()

```

```{r}

#res = model; dat = data

forest(Genotype_basemod, atransf = exp, at=log(c(.05, .25, 1, 4,10,55)), xlim=c(-12,10), cex=.75, header="Study ID", mlab="", efac = 2, order = "obs"       )

### add text with Q-value, dfs, p-value, and I^2 statistic
text(-12, -1, pos=4, cex=0.75, bquote(paste("REM (Q = ",
     .(formatC(Genotype_basemod$QE, digits=2, format="f")), ", df = ", .(Genotype_basemod$k - Genotype_basemod$p),
     ", p = ", .(formatC(Genotype_basemod$QEp, digits=2, format="f")), "; ", I^2, " = ",
     .(formatC(I2, digits=1, format="f")), "%)")))


```

**Conclusion:** Overall Risk Ratio Estimate:  2.31
Mortality is consistently higher in the "experimental" group, i.e. in panzootic/GPL Genotypes


**Mortality Plot**
```{r}

Genotype_RR_data_Mortality = Genotype_RR_data %>%
  mutate(mortality.endemicCOR = mortality.endemic-Amount.mortality,mortality.pandemicCOR = mortality.pandemic-Amount.mortality)%>%
  pivot_longer(cols = mortality.endemicCOR:mortality.pandemicCOR)%>%
  filter(!name=="Nendemic") %>%
  arrange(desc(mortality.pandemic))%>%
  mutate(Corrected.Mortality = if_else(value < 0, 0, value))

#Genotype_RR_data_Mortality %>%
 # ggplot(aes(x= Study.Name, y = value)) +
  #geom_point(aes(color = name, alpha = .5)) +
  #theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  #ylab("corrected Mortality (%)") +
  #facet_wrap(~Historical.coexistence,scales = "free")


Genotype_RR_data_Mortality %>%
  ggplot(aes(fill=name, y=Corrected.Mortality, x=Study.Name)) + 
    geom_bar(position="dodge", stat="identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ylab("corrected mortality (%)") +
  facet_wrap(~Historical.coexistence,scales = "free")+
  ggthemes::scale_fill_colorblind()


Genotype_RR_data_Mortality %>%
  ggplot(aes(y=Corrected.Mortality, x=name, fill = name)) + 
    geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ylab("corrected mortality (%)") +
  facet_wrap(~Historical.coexistence,scales = "free")+
  ggthemes::scale_fill_colorblind()

```

**Comparison of Variances between Historical coexistence categories**

**Conclusion**:  Not Significant

```{r}
Genotype_RR_data_Mortality_Pan = Genotype_RR_data_Mortality %>%
  filter(name=="mortality.pandemicCOR")

fligner.test(Corrected.Mortality ~ Historical.coexistence, data = Genotype_RR_data_Mortality_Pan)

Genotype_RR_data_Mortality_End = Genotype_RR_data_Mortality %>%
  filter(name=="mortality.endemicCOR")

fligner.test(Corrected.Mortality ~ Historical.coexistence, data = Genotype_RR_data_Mortality_End)
```


## Detecting Outliers

```{r}
#influential > 0.45
Genotype_CD = cooks.distance.rma.mv(Genotype_basemod)

plot(Genotype_CD, type="o", pch=19, xlab="Observed Outcome", ylab="Cook's Distance")

#influential = >1
Genotype_dfB = dfbetas.rma.mv(Genotype_basemod)

Genotype_dfB

# Hat value outliers < 3*(1/k) == 3*(1/16) == 0.167  
Genotype_Hat = hatvalues.rma.mv(Genotype_basemod)

plot(Genotype_Hat, type="o", pch=19, xlab="Observed Outcome", ylab="Hat value")
```

**Outliers:** None

## SubGroup/Moderator models

Useful for interpretation: https://www.metafor-project.org/doku.php/tips:meta_regression_with_log_rr

### Multi-factor model

```{r}

Genotype_basemod_noOL_Full = rma.mv(yi=yi, V=vi,data = Genotype_RR_data, random = ~1|Shortcitation, mods = ~ host.adult.ecology  + life.stage.exp + Historical.coexistence,slab = Genotype_RR_data$Study.Name, method = "REML")

## model results
summary(Genotype_basemod_noOL_Full)
```

**model comparison**

```{r}
Genotype_basemodML = rma.mv(yi=yi, V=vi,data = Genotype_RR_data, random = ~1|Shortcitation,slab = Genotype_RR_data$Study.Name, method = "ML")

Genotype_basemod_FullML = rma.mv(yi=yi, V=vi,data = Genotype_RR_data, random = ~1|Shortcitation, mods = ~ host.adult.ecology  + life.stage.exp + Historical.coexistence,slab = Genotype_RR_data$Study.Name, method = "ML")

## says REML is not meaningful?  Switched to ML
anova(Genotype_basemodML,Genotype_basemod_FullML)
```

**Conclusion** Full model is "better" than intercept only model 

*random study term included in both*

### Single factor models

**Host Ecology**

Conclusion: Not Significant

```{r}

Genotype_basemod_noOL_HEco = rma.mv(yi=yi, V=vi,data = Genotype_RR_data, random = ~1|Shortcitation, mods = ~ host.adult.ecology, slab = Genotype_RR_data$Study.Name, method = "REML")

## model results
summary(Genotype_basemod_noOL_HEco)

```

**Life Stage**

Conclusion: Significant, metamorph category

```{r}
Genotype_RR_data2 = Genotype_RR_data %>%
  arrange(life.stage.exp)
Genotype_basemod_noOL_LS = rma.mv(yi=yi, V=vi,data = Genotype_RR_data2, random = ~1|Shortcitation, mods = ~ + life.stage.exp,slab = Genotype_RR_data2$Study.Name, method = "REML")

## model results
summary(Genotype_basemod_noOL_LS)

```

**Sub-group Plot**

*the formatting code for this is kind of ridiculous so still working on this....*

*edited inn illustrator for manuscript*


```{r}

### a little helper function to add Q-test, I^2, and tau^2 estimate info
mlabfun <- function(text, Genotype_basemod_noOL_LS) {
   list(bquote(paste(.(text),
      " (Q = ", .(formatC(Genotype_basemod_noOL_LS$QE, digits=2, format="f")),
      ", df = ", .(Genotype_basemod_noOL_LS$k - Genotype_basemod_noOL_LS$p),
      ", p ", .(metafor:::.pval(Genotype_basemod_noOL_LS$QEp, digits=2, showeq=TRUE, sep=" ")))))}
 
### set up forest plot (with 2x2 table counts added; the 'rows' argument is
### used to specify in which rows the outcomes will be plotted)
forest(Genotype_basemod_noOL_LS, xlim=c(-12, 10), at=log(c(0.05, 0.25, 1, 4,10,55)), atransf=exp,
      cex=0.6, ylim=c(-1, 28),
      rows=c(3:9,12:16,19:21),
       mlab=mlabfun("RE Model for All Studies", Genotype_basemod_noOL_LS),
       psize=1, header="Study ID")
op <- par(cex=0.75, font=2)
### add text for the subgroups
text(-12, c(22,17,10), pos=4, c("Metamorph",
                               "Larvae*",
                               "Adult"))
 
### set par back to the original settings
par(op)
 
### fit random-effects model in the three subgroups
Genotype_basemod_noOL_LSa = rma.mv(yi=yi, V=vi,data = Genotype_RR_data, random = ~1|Shortcitation,slab = Genotype_RR_data$Study.Name, method = "REML",subset=(life.stage.exp=="adult"))
Genotype_basemod_noOL_LSl = rma.mv(yi=yi, V=vi,data = Genotype_RR_data, random = ~1|Shortcitation,slab = Genotype_RR_data$Study.Name, method = "REML",subset=(life.stage.exp=="exposed as larvae, allowed to metamorphose"))
Genotype_basemod_noOL_LSm = rma.mv(yi=yi, V=vi,data = Genotype_RR_data, random = ~1|Shortcitation,slab = Genotype_RR_data$Study.Name, method = "REML",subset=(life.stage.exp=="metamorph"))
 
### add summary polygons for the three subgroups
addpoly(Genotype_basemod_noOL_LSa, row=2, cex=0.6, atransf=exp, mlab=mlabfun("REM", Genotype_basemod_noOL_LSa))
addpoly(Genotype_basemod_noOL_LSl, row= 11, cex=0.6, atransf=exp, mlab=mlabfun("REM", Genotype_basemod_noOL_LSl))
addpoly(Genotype_basemod_noOL_LSm, row= 18, cex=0.6, atransf=exp, mlab=mlabfun("REM", Genotype_basemod_noOL_LSm))
 
### add text for the test of subgroup differences
text(-12, -1.8, pos=4, cex=0.6, bquote(paste("Test for Subgroup Differences: ",
     Q[M], " = ", .(formatC(Genotype_basemod_noOL_LS$QM, digits=2, format="f")), ", df = ", .(Genotype_basemod_noOL_LS$p - 1),
     ", p = ", .(formatC(Genotype_basemod_noOL_LS$QMp, digits=2, format="f")))))

```

**Effects for Subgroups**

Risk Ratio for Adults

```{r}
exp(coef(Genotype_basemod_noOL_LS)[[1]])

predict(Genotype_basemod_noOL_LS, newmods=c(0,0), transf=exp, digits=2)
```

Risk Ratio of Larvae*

```{r}
exp(coef(Genotype_basemod_noOL_LS)[[1]] + coef(Genotype_basemod_noOL_LS)[[2]])
```

How much greater is larval risk vs adult?

```{r}
exp(coef(Genotype_basemod_noOL_LS)[[2]])

```

Risk Ratio of Metamorph

```{r}

exp(coef(Genotype_basemod_noOL_LS)[[1]] + coef(Genotype_basemod_noOL_LS)[[3]])
```

How much greater is metamorph risk vs adult?

```{r}
exp(coef(Genotype_basemod_noOL_LS)[[3]])
```

How much greater is metamorph risk vs larvae?

```{r}
(exp(coef(Genotype_basemod_noOL_LS)[[1]] + coef(Genotype_basemod_noOL_LS)[[3]])) / (exp(coef(Genotype_basemod_noOL_LS)[[1]] + coef(Genotype_basemod_noOL_LS)[[2]]))

```


**Historical Coexistence**

Conclusion: **NOT** Significant

```{r}
Genotype_RR_data3 = Genotype_RR_data %>%
  arrange(Historical.coexistence)
Genotype_basemod_noOL_HisCo = rma.mv(yi=yi, V=vi,data = Genotype_RR_data3, random = ~1|Shortcitation, mods = ~ Historical.coexistence,slab = Genotype_RR_data3$Study.Name, method = "REML")

## model results
summary(Genotype_basemod_noOL_HisCo)

```

# **Local Adaptation Hypothesis**

## Main model

```{r}

## using event rate data

## we need number of individuals data...recalculating percantages to numbers 
MetaBdData_Location2 = MetaBdData_Location %>%
  mutate(nonLocal.ndead = round((mortality.nonlocalBd-Amount.mortality)/100*nnonlocalBd),
         Local.ndead=round((mortality.localBd-Amount.mortality)/100*nlocalBd)) %>%
  mutate(nonLocal.nlive = nnonlocalBd-nonLocal.ndead,Local.nlive = nlocalBd-Local.ndead) %>%
  mutate(Local.ndead.cor = if_else(Local.ndead < 0, 0, Local.ndead)) %>%
  mutate(nonLocal.ndead.cor = if_else(nonLocal.ndead < 0, 0, nonLocal.ndead))

## calculating the risk ratios from the raw # individuals data
Location_RR_data = metafor::escalc(measure="RR", ai=nonLocal.ndead.cor, bi=nonLocal.nlive, ci=Local.ndead.cor, di=Local.nlive, data=MetaBdData_Location2)

## main model with a random term for study 
Location_RR_data_GPLs = Location_RR_data %>%
  filter(Location.GPLGPL== "Y")

Location_basemod_GPLs = rma.mv(yi=yi, V=vi,data = Location_RR_data_GPLs, random = ~1|Shortcitation,slab = Location_RR_data_GPLs$Study.Name, method = "REML")

## model results
summary(Location_basemod_GPLs)

```

## Heterogeneity: I^2

```{r}
# from metafor package site
W <- diag(1/Location_RR_data_GPLs$vi)
X <- model.matrix(Location_basemod_GPLs)
P <- W - W %*% X %*% solve(t(X) %*% W %*% X) %*% t(X) %*% W
I2loc = 100 * sum(Location_basemod_GPLs$sigma2) / (sum(Location_basemod_GPLs$sigma2) + (Location_basemod_GPLs$k-Location_basemod_GPLs$p)/sum(diag(P)))
I2loc
```

**Overall all risk ratio**

*estimate in model output is log RR*

```{r}
## converting log estimate to RR
exp(coef(Location_basemod_GPLs)[[1]])

## https://www.metafor-project.org/doku.php/tips:meta_regression_with_log_rr

```

## Plots

```{r}

#res = model; dat = data

forest(Location_basemod_GPLs, atransf=exp, at=log(c(.05, .25, 1, 4,10)), xlim=c(-16,10), cex=.75, header="Study ID", mlab="")

### add text with Q-value, dfs, p-value, and I^2 statistic
text(-16, -1, pos=4, cex=0.75, bquote(paste("RE Model (Q = ",
     .(formatC(Location_basemod_GPLs$QE, digits=2, format="f")), ", df = ", .(Location_basemod_GPLs$k - Location_basemod_GPLs$p),
     ", p = ", .(formatC(Location_basemod_GPLs$QEp, digits=2, format="f")), "; ", I^2, " = ",
     .(formatC(I2loc, digits=1, format="f")), "%)")))

```

**Conclusion**: model not significant, no overarching effect in a given direction

## Outliers

```{r}
Location_CD = cooks.distance.rma.mv(Location_basemod_GPLs)

plot(Location_CD, type="o", pch=19, xlab="Observed Outcome", ylab="Cook's Distance")

Location_CD %>% as.data.frame()

#influential = >1
Location_dfB = dfbetas.rma.mv(Location_basemod_GPLs)

Location_dfB

# Hat value outliers > 3*(1/k) == 3*(1/17) == 0.158 
Location_Hat = hatvalues.rma.mv(Location_basemod_GPLs)

plot(Location_Hat, type="o", pch=19, xlab="Observed Outcome", ylab="Hat value")
  
```

Outliers: None

## SubGroup/Moderator models

Useful for interpretation: https://www.metafor-project.org/doku.php/tips:meta_regression_with_log_rr

**multi-factor model**

```{r}

Location_basemod_GPLs_Full = rma.mv(yi=yi, V=vi,data = Location_RR_data_GPLs, random = ~1|Shortcitation,mods = ~ host.adult.ecology  + life.stage.exp, slab = Location_RR_data_GPLs$Study.Name, method = "REML")

summary(Location_basemod_GPLs_Full)

```

**model comparison of GPL-GPL only data**

```{r}
Location_basemod_GPLs_ML = rma.mv(yi=yi, V=vi,data = Location_RR_data_GPLs, random = ~1|Shortcitation,slab = Location_RR_data_GPLs$Study.Name, method = "ML")

Location_basemod_GPLs_Full_ML = rma.mv(yi=yi, V=vi,data = Location_RR_data_GPLs, random = ~1|Shortcitation,mods = ~ host.adult.ecology  + life.stage.exp, slab = Location_RR_data_GPLs$Study.Name, method = "ML")

anova(Location_basemod_GPLs_ML,Location_basemod_GPLs_Full_ML)
```

**Conclusion** Full model is "NOT better" than intercept only model 
*random study term included in both*


**single factor models**

*maybe exclude because the full model isn't better than base model*

**Host Ecology**

Conclusion: Not Significant

```{r}

Location_basemod_noOL_HEco = rma.mv(yi=yi, V=vi,data = Location_RR_data_GPLs, random = ~1|Shortcitation, mods = ~ host.adult.ecology, slab = Location_RR_data_GPLs$Study.Name, method = "REML")

## model results
summary(Location_basemod_noOL_HEco)

```

**Life Stage**

Conclusion: Not Significant

```{r}

Location_basemod_noOL_LS = rma.mv(yi=yi, V=vi,data = Location_RR_data_GPLs, random = ~1|Shortcitation, mods = ~ + life.stage.exp,slab = Location_RR_data_GPLs$Study.Name, method = "REML")

## model results
summary(Location_basemod_noOL_LS)

```
